---
title: "Assignment 1"
author: "Khan Safayet Hossin"
date: "Wednesday, May 13, 2015"
output: html_document
---
In this assignment some exploratory analysis has been done using <a href="https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"> Activity monitoring dataset</a>[52K] dataset.

###Loading data
There are 3 variables in the dataset:

- <b>steps</b>: Number of steps taking in a 5-minute interval (missing values are coded as `NA`)
- <b>date</b>: The date on which the measurement was taken in YYYY-MM-DD format
- <b>interval</b>: Identifier for the 5-minute interval in which measurement was taken

As a first step data is loaded into a data frame in R and some basic checks are performed.

```{r}
data.r <- read.csv("activity.csv", header=TRUE, sep=",", as.is=TRUE)
nrow(data.r)
summary(data.r)
```
###Missing values and analysis dataset preparation
Missing values are checked in the loaded dataset
```{r}
colSums(is.na(data.r))
```
Two datasets are prepared for the analysis:

- Excluding any missing data in steps 
- Replacing missing data in steps with the median steps across all days and all intervals
```{r}
## dataset 1
data.f1 <- data.r[!is.na(data.r$steps),]
all(colSums(is.na(data.f1))==0)
## dataset 2
data.f2 <- as.data.frame(data.r)
data.f2[is.na(data.f2$steps),1] <- median(data.f2$steps, na.rm=TRUE)
all(colSums(is.na(data.f2))==0)
```

###Steps taken per day

Daily steps summary (total steps per day, median steps per day, mean steps per day) has been calculated for this part of analysis: 
```{r}
## dataset 1
sm <-aggregate(data.f1$steps, by=list(data.f1$date), function(x){c(sum(x), median(x), mean(x))})
data.s1<- cbind(as.data.frame(sm$Group), sm$x)
names(data.s1) <- c("date", "totSteps", "medSteps", "meaSteps")
data.s1
## dataset 2
sm <-aggregate(data.f2$steps, by=list(data.f2$date), function(x){c(sum(x), median(x), mean(x))})
data.s2<- cbind(as.data.frame(sm$Group), sm$x)
names(data.s2) <- c("date", "totSteps", "medSteps", "meaSteps")
data.s2
```

Next, the histogram of total number of steps taken each day is generated:

```{r}
##png(filename="plot1.png")
par(mfrow=c(1,2))
## dataset 1
hist(data.s1$totSteps, breaks=20, xlab = "Total steps a day", main = "Excluding NAs", col="red")
## dataset 2
hist(data.s2$totSteps, breaks=20, xlab = "Total steps a day", main = "Including NAs", col="blue")
##dev.off()
```

Also, the Box plot of total number of steps taken each day highlighting the mean as red horizontal line.
```{r}
m1 <-round(mean(data.s1$totSteps),0)
m2 <-round(mean(data.s2$totSteps),0)
##png(filename="plot2.png")
par(mfrow=c(1,2))
## dataset 1
boxplot(data.s1$totSteps, ylab = paste("Total steps a day (mean:", m1, ")"), main = "Excluding NAs")
abline(h=mean(data.s1$totSteps),col="red", lwd=2)
## dataset 2
boxplot(data.s2$totSteps, ylab = paste("Total steps a day (mean:",m2,")"), main = "Including NAs")
abline(h=mean(data.s2$totSteps),col="red", lwd=2)
##dev.off()
```

It is seen that the difference in the spread of data across the time is not that signification including/excluding the `NA's`, however, the mean has changed about a thousand steps.

###Average daily activity

Average daily activity pattern is analyzed including and excluding `NA's` using the time series plot. 
```{r}
## dataset 1
sm1 <-aggregate(data.f1$steps, by=list(data.f1$interval), mean)
maxStepInterval1 <- sm1[sm1["x"]==max(sm1["x"]),1]
r1 <- range(sm1$x)
## dataset 2
sm2 <-aggregate(data.f2$steps, by=list(data.f2$interval), mean)
maxStepInterval2 <- sm2[sm2["x"]==max(sm2["x"]),1]
r2 <- range(sm2$x)
y.rng <-c(min(r1, r2), max(r1,r2))

##png(filename="plot3.png")
par(mfrow=c(1,1))
plot(sm1$Group, sm1$x, type="l", col="blue", 
     xlab= paste("intervals (max excluding NAs:", maxStepInterval1, ", max excluding NAs:", maxStepInterval2,")"), 
     main="Average daily activity", ylab="steps in 5 minues", ylim=y.rng)
lines(sm2$Group, sm2$x, col="red", type="l")
abline(v=sm1[sm1["x"]==max(sm1["x"]),1], col="black", lwd=2)
abline(v=sm2[sm2["x"]==max(sm2["x"]),1], col="green", lwd=2)
legend("topright", legend=c("excluding NAs","including NAs"), col = c("blue","red"), ncol = 1, lwd = 1)
##dev.off()
```

It is seen in the time series plot that the patterns have not changed significantly due to inclusion and exclusion of the `NA's` in the dataset. The maximum steps interval remain same for both at 8:35 AM which is shown as green line in both plots.

###Weekdays/Weekends activity pattern

Average activity pattern during weekdays and weekends is also analyzed excluding `NA's` only as no difference has been observed in average daily actipity pattern including and excluding `NA's`.

```{r}
## dataset 1
##png(filename="plot4.png")
data.f1$dateT <- strptime(data.f1$date, "%Y-%m-%d") 
data.f1$dateType <- weekdays(data.f1$dateT, abbreviate=TRUE)
data.f1[data.f1$dateType %in% c("Sat","Sun"),5]<- "weekend"
data.f1[data.f1$dateType %in% c("Mon","Tue", "Wed","Thu","Fri"),5]<- "weekday"
data.f1$dateType <- as.factor(data.f1$dateType)
sm3 <-aggregate(data.f1[data.f1$dateType=="weekday",1], by=list(data.f1[data.f1$dateType=="weekday",3]), mean)
data.s3 <- cbind(as.data.frame("weekday"), sm3$Group, sm3$x)
names(data.s3)<- c("dateType", "interval", "steps")

sm4 <-aggregate(data.f1[data.f1$dateType=="weekend",1], by=list(data.f1[data.f1$dateType=="weekend",3]), mean)
data.s4 <-cbind(as.data.frame("weekend"), sm4$Group, sm4$x)
names(data.s4)<- c("dateType", "interval", "steps")

data.s5<- rbind(data.s3, data.s4)

library(lattice)
xyplot(data.s5$steps~data.s5$interval | data.s5$dateType, type="l", 
       xlab = "Interval", ylab = "Number of steps",  layout=c(1,2))
##dev.off()
```

A pattern of higher number of steps in the morning is seen during weekdays(5:00 AM to 9:30 AM), where in weekends a pattern of higher number of steps is seen during the rest of the day(9:30 AM to 8:00 PM) in the weekend/weekday activity plot.
